A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-09-18 manuscript/manuscript.Rmd
update manuscript

---
title             : "A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data"
shorttitle        : "Tutorial on hazard analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. The orthodox method for analysing such data -- comparing means with analysis-of-variance -- is actually hiding a lot of information about psychological effects, such as their onset time and duration, and whether they are time-locked to stimuli. Such information can change key conclusions about psychological processes and can be revealed by using distributional measures.  
  Here we provide a set of tutorials on how to implement one particular distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. We illustrate how one can calculate the descriptive statistics, and how one can implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. The R code is publicly available on Github and OSF, and can easily be adapted for other data sets. We further discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time hazard analysis.

@@

In this final tutorial we illustrate how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of @panisWhatShapingRT2016.

@@

@@

- role of response deadlines, low-level vs. higher-level processes, 
- clustering algorithms based on h(t) and ca(t) data
-

>>>>

---
title             : "A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data"
shorttitle        : "Tutorial on hazard analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. The orthodox method for analysing such data -- comparing means with analysis-of-variance -- is actually hiding a lot of information about psychological effects, such as their onset time and duration, and whether they are time-locked to stimuli. Such information can change key conclusions about psychological processes and can be revealed by using distributional measures.  
  Here we provide a set of tutorials on how to implement one particular distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. We illustrate how one can calculate the descriptive statistics, and how one can implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. The R code is publicly available on Github and OSF, and can easily be adapted for other data sets. We  discuss possible link functions and prior distributions, how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time hazard analysis.

@@

In this final tutorial we illustrate how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of @panisWhatShapingRT2016, which includes two independent variables: mask type and prime type.

@@

We noticed that many researchers are still reluctant to abandon analysis-of-variance and switch to event history analysis when analysing time-to-event data. By providing this tutorial, we hope that researchers will start using hazard analysis more often.
While we focused on within-subject factorial small-*N* designs, event history analysis has been developed to also encompass other designs.

## Advantages of hazard functions

@@

- role of response deadlines, low-level vs. higher-level processes, overlap between distributions
- clustering algorithms based on h(t) and ca(t) data
-

### 2024-09-18 manuscript/manuscript.tex
update manuscript

\abstract{%
Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. The orthodox method for analysing such data -- comparing means with analysis-of-variance -- is actually hiding a lot of information about psychological effects, such as their onset time and duration, and whether they are time-locked to stimuli. Such information can change key conclusions about psychological processes and can be revealed by using distributional measures.\\
Here we provide a set of tutorials on how to implement one particular distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. We illustrate how one can calculate the descriptive statistics, and how one can implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. The R code is publicly available on Github and OSF, and can easily be adapted for other data sets. We further discuss how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time hazard analysis.
}

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

In this final tutorial we illustrate how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of Panis and Schmidt (2016).

@@

@@

\begin{itemize}
\tightlist
\item
  role of response deadlines, low-level vs.~higher-level processes,
\item
  clustering algorithms based on h(t) and ca(t) data
\item
\end{itemize}

>>>>

\abstract{%
Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. The orthodox method for analysing such data -- comparing means with analysis-of-variance -- is actually hiding a lot of information about psychological effects, such as their onset time and duration, and whether they are time-locked to stimuli. Such information can change key conclusions about psychological processes and can be revealed by using distributional measures.\\
Here we provide a set of tutorials on how to implement one particular distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. We illustrate how one can calculate the descriptive statistics, and how one can implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. The R code is publicly available on Github and OSF, and can easily be adapted for other data sets. We discuss possible link functions and prior distributions, how to manage inter-individual differences, implications for experimental design, and how to select among various options when analysing time-to-event data using discrete-time hazard analysis.
}

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

In this final tutorial we illustrate how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of Panis and Schmidt (2016), which includes two independent variables: mask type and prime type.

@@

We noticed that many researchers are still reluctant to abandon analysis-of-variance and switch to event history analysis when analysing time-to-event data. By providing this tutorial, we hope that researchers will start using hazard analysis more often.
While we focused on within-subject factorial small-\emph{N} designs, event history analysis has been developed to also encompass other designs.

\subsection{Advantages of hazard functions}\label{advantages-of-hazard-functions}

@@

\begin{itemize}
\tightlist
\item
  role of response deadlines, low-level vs.~higher-level processes, overlap between distributions
\item
  clustering algorithms based on h(t) and ca(t) data
\item
\end{itemize}
